---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1

========================================

## Loading and preprocessing the data

First of all we need to load the data into a variable

```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
data$date <- as.Date(data$date)
```

Our data looks like this:

```{r}
summary(data)
```

I also want to operate data via dplyr sintax, so I will load dplyr package and 
perform a transformation to dplyr data format.

```{r}
library(dplyr)
d <- tbl_df(data)
d
rm(data)
```

## What is mean total number of steps taken per day?

Let's calculate total number of steps taken per day.

```{r}
steps_per_day <- d %>% group_by(date) %>% summarize(steps_sum = sum(steps, na.rm=TRUE))
```

Let's make a histogram of the total number of steps taken each day

```{r}
hist(steps_per_day$steps_sum, 
     breaks = 16, 
     main = "Histogram of steps per day", 
     xlab = "Sum of steps per day",
     col = "grey")
```

Calculate and report the mean and median of the total number of steps taken per day.

```{r}
avg <- mean(steps_per_day$steps_sum)
#9354.23
med <- median(steps_per_day$steps_sum)
#10395
```

Mean is `r avg` and median is `r med`.

## What is the average daily activity pattern?

Let's calculate average number of steps taken per each 5-minute time interval and plot it.

```{r}
steps_avg_interval <- d %>% group_by(interval) %>% summarize(steps_avg = mean(steps, na.rm=TRUE))
plot(steps_avg_interval, 
     type = "l", 
     main = "Daily activity pattern", 
     xlab = "5-minute Interval", 
     ylab = "Average steps taken")

```

Then find a 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.

```{r}
max_steps <- max(steps_avg_interval$steps_avg)
max_interval <- as.numeric(steps_avg_interval %>% 
                               filter(steps_avg == max_steps) %>% 
                               select(interval)
                           )
# 206.1698
# 835
```

The maximum average number of steps is `r max_steps` and it was taken in `r max_interval` interval.

## Imputing missing values

Lets' calculate the total number of missing values in our dataset.

```{r}
cnt_na <- sum(is.na(d$steps))
```

Total number of missing (NA) values is `r cnt_na` (it also can be seen in first data summary).

I'll fill in all missing values with mean value for corresponding 5-minute interval. 
Remember that these mean values were calculated during previous step.

For these manipulations i'll create a new dataset equal to original dataset so as to make transformation on it.

```{r}
d2 <- d
merged <- tbl_df(merge(d2, steps_avg_interval, by.x = "interval", by.y = "interval", all.x = TRUE)) %>%
    mutate(steps_new = steps)

indexes <- which(is.na(merged$steps)) 
merged$steps_new[ indexes ] <- merged$steps_avg[ indexes ]
```

Make a histogram of the total number of steps taken each day and 

```{r}
steps_per_day1 <- merged %>% group_by(date) %>% summarize(steps_sum = sum(steps_new))
hist(steps_per_day1$steps_sum, 
     breaks = 16,
     main = "Histogram of steps per day\n(missing values were replaced)", 
     xlab = "Sum of steps per day",
     col = "pink")
```

```{r}
avg1 <- sprintf("%.2f", mean(steps_per_day1$steps_sum))
# 10766.19
med1 <- sprintf("%.2f", median(steps_per_day1$steps_sum))
# 10766.19
```

Mean is `r avg1` and median is `r med1`.

As we can see new values of mean and median differ from previously calculated estimated values:
- mean: `r avg1` instead of `r avg`;
- median: `r med1` instead of `r med`.
The impact of imputing missing data is enlarging values.

## Are there differences in activity patterns between weekdays and weekends?

For data manipulating I'll use wday() function from lubridate package 
(instead of regular weekdays() function).

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
library(lubridate)
merged <- merged %>% mutate(wday = wday(date), dtype = "Weekday")
# days 1 and 7 are weekend days
indexes1 <- which(merged$wday %in% c(1,7))
merged$dtype[ indexes1 ] <- "Weekend"
merged %>% mutate(dtype = as.factor(dtype))
```

Let's calculate activity patterns for weekdays and weekends 

```{r}
steps_avg_dtype <- merged %>% 
    group_by(dtype, interval) %>% 
    summarize(steps_avg = mean(steps_new)) %>%
    print()

```

and then make a panel plot to show differences between them.

```{r}
library(lattice)
xyplot(steps_avg ~ interval | dtype, 
       data=steps_avg_dtype, 
       layout = c(1,2), 
       type = "l", 
       xlab = "Interval", 
       ylab = "Number of steps", 
       main = "Activity patterns for weekdays and weekends ")
```

As you can see, the weekend distribution of steps is more spread out over the time intervals than the weekdays distribution.